\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-10-06 lang-design.tex
stuff

Crucially, the types of the constructors are polymorphic to their enclosing family---%
their meaning is \emph{late bound}.
For example, although "tm_app" is defined in the base family,
it can be used to construct applications of fixpoints in the derived family:
%such as \lstinline[mathescape]{tm_app (tm_fix "f" (tm_abs "x" $\dadada$)) tm_unit},
its type \lsti{tm -> tm -> tm} says "tm_app" can be applied to any
Gallina terms of type "tm",
which in family "STLC_fix" include those constructed from "tm_fix".

@@

\begingroup
\newcommand\hlcase[1]{\tikzmarkin[disable rounded corners=true,set fill color=yellow!18,set border color=yellow!18]{#1}(0.00,-0.05)(-0.00,0.20)}
\begin{lstlisting}[basicstyle=\fontsize{8}{9}\ttfamily]
\forall (P : tm -> Type), [@\hlcase{tm-recursor-case-var}@](\forall x, P (tm_var x))[@\tikzmarkend{tm-recursor-case-var}@] -> [@\hlcase{tm-recursor-case-unit}@]P tm_unit[@\tikzmarkend{tm-recursor-case-unit}@] -> [@\hlcase{tm-recursor-case-abs}@](\forall x t, P t -> P (tm_abs x t))[@\tikzmarkend{tm-recursor-case-abs}@] ->
  [@\hlcase{tm-recursor-case-app}@](\forall t1 t2, P t1 -> P t2 -> P (tm_app t1 t2))[@\tikzmarkend{tm-recursor-case-app}@] -> \forall t, P t
\end{lstlisting}
\endgroup

@@

As an example, consider the substitution function "subst", defined using
\lsti!FRecursion!.
The \lsti{on} clause specifies that recursion is over "tm".
The \lsti{motive} clause suggests that the recursive function being
defined has type \lsti{tm -> id -> tm -> tm}.
%
The "subst" function in "STLC_fix" further binds that in "STLC":
the four cases from "STLC" are automatically inherited and reused,
with "STLC_fix" retroactively adding a fifth case to form the new "subst" function.
%
For exhaustivity, it is a static error if the programmer fails to
further bind "subst" and define this fifth case.
The type system does this check by examining if the inductive type "tm",
over which "subst" is recursively defined, is further bound in the same family.

Like every other field of the family, the meaning of "subst" is late bound.
For example, the type of "st_beta" (modeling the $\beta$-reduction rule)
in family "STLC" refers to "subst".
When "st_beta" is inherited into family "STLC_fix", it has a type that now
refers to the "subst" function in the same family, where "subst" is defined
by handling all the five cases known to that family.

The \lstinline{FInduction} command is similar to \lstinline{FRecursion} but
allows cases to be defined in proof mode.
For example, "weakenlem" is proven by induction on the typing relation
"hasty".
Its \lsti{motive} clause suggests that the lemma reads as
\lsti{\forall G t T, hasty G t T -> \forall G', includedin G G' -> hasty G' t T}.
Upon entering case "ht_abs", for instance, Coq enters proof mode with the goal

@@

\noindent
The programmer can discharge the goal using tactic programming.
The idiomatic programming experience with Coq is largely preserved.

>>>>

Crucially, the types of the constructors are polymorphic to their enclosing family---%
their meaning is \emph{late bound}.
For example, although "tm_app" is defined in the base family,
it can be used to construct applications of fixpoints in the derived family:
%such as \lsti[mathescape]{tm_app (tm_fix "f" (tm_abs "x" $\dadada$)) tm_unit},
its type \lsti{tm -> tm -> tm} says "tm_app" can be applied to any
Gallina terms of type "tm",
which in family "STLC_fix" include those constructed from "tm_fix".

@@

\begingroup
\newcommand\hlcase[1]{\tikzmarkin[disable rounded corners=true,set fill color=yellow!18,set border color=yellow!18]{#1}(0.00,-0.05)(-0.00,0.20)}
\begin{lstlisting}[basicstyle=\fontsize{8.25}{9}\ttfamily]
\forall (P : tm -> Type), [@\hlcase{tm-recursor-case-var}@](\forall x, P (tm_var x))[@\tikzmarkend{tm-recursor-case-var}@] -> [@\hlcase{tm-recursor-case-unit}@]P tm_unit[@\tikzmarkend{tm-recursor-case-unit}@] -> [@\hlcase{tm-recursor-case-abs}@](\forall x t, P t -> P (tm_abs x t))[@\tikzmarkend{tm-recursor-case-abs}@] ->
  [@\hlcase{tm-recursor-case-app}@](\forall t1 t2, P t1 -> P t2 -> P (tm_app t1 t2))[@\tikzmarkend{tm-recursor-case-app}@] -> \forall t, P t
\end{lstlisting}
\endgroup

@@

As an example, consider the substitution function "subst", defined using
\lsti!FRecursion!.
The \lsti{on} clause specifies that recursion is over "tm".
The \lsti{motive} clause suggests that the recursive function being
defined has type \lsti{tm -> id -> tm -> tm}.
%
The "subst" function in "STLC" is further bound by the "subst" in "STLC_fix":
the four cases from "STLC" are automatically inherited and reused,
with "STLC_fix" retroactively adding a fifth case to form the new "subst" function.
%
For exhaustivity, it is a static error if the programmer fails to
further bind "subst" and define this fifth case.
The type system does this check by examining if the inductive type "tm",
over which "subst" is recursively defined, is further bound in the same family.

Like all references to the fields of a family, the meaning of "subst" is late bound.
For example, the type of "st_beta" (modeling the $\beta$-reduction rule)
in family "STLC" refers to "subst".
When "st_beta" is inherited into family "STLC_fix", it has a type that now
refers to the "subst" function in "STLC_fix", where "subst" is defined
by handling all the five cases known to that family.

The \lsti{FInduction} command is similar to \lsti{FRecursion} but
allows cases to be defined in proof mode.
For example, "weakenlem" is proven by induction on the typing relation
"hasty".
Its \lsti{motive} clause suggests that the lemma reads as
\lsti{\forall G t T, hasty G t T -> \forall G', includedin G G' -> hasty G' t T}.
Upon entering case "ht_abs", for instance, Coq enters proof mode with the goal

@@

\noindent
The programmer can use tactic programming to discharge the goal.
%The idiomatic programming experience with Coq is largely preserved.

\noindentparagraph{Equality on late bound definitions.}

Consider proving the "ht_unit" case of the substitution lemma, "substlem".
The goal is seemingly trivial: the programmer is asked to prove

\begin{lstlisting}[basicstyle=\fontsize{8.25}{9}\ttfamily]
\forall G' G x t' T', G' = extend G x T' -> hasty empty t' T' -> hasty G (subst tm_unit x t') ty_unit.
\end{lstlisting}

\noindent
If "subst" were an ordinary Coq function, then the programmer could
discharge the goal with
\lsti{intros; simpl subst; apply ht_unit.}
The Ltac term \lsti{simpl subst} unfolds \lsti{subst} using its definition
and simplifies \lsti{subst tm_unit x t'} to \lsti{tm_unit}.

However, with \lsti{subst} being late bound, \lsti{subst} can no longer
be unfolded: the definition of \lsti{subst}, as a recursive function,
varies across families, yet the proof of the \lsti{ht_unit} case should be
reusable by a derived family regardless of how that family may add
new cases to the definition of \lsti{subst}.
How can the programmer change the goal to \lsti{hasty G tm_unit ty_unit}, then?

A key observation is that although late binding
prevents \emph{definitional equality} on \lsti{subst}, it does not
affect \emph{propositional equality}.
That is, \lsti{subst tm_unit x t' = tm_unit}, as a proposition
(\lsti{Prop}), can still be proved.
After all, how \lsti{subst} is defined on \lsti{tm_unit} does not vary
from a base family to a derived family.
%
Hence, our plugin provides a tactic \lsti{fsimpl}
with which the programmer can rewrite \lsti{subst tm_unit x t'} to
\lsti{tm_unit}.
